---
title: "Lesson 1: Tidy Data Manipulation I"
author: "David Robinson"
date: "January 26, 2015"
output: pdf_document
---

```{r echo = FALSE}
library(knitr)
opts_chunk$set(cache = TRUE, message = FALSE, warning = FALSE)
```

### Setup and Installation

First you need to install the dplyr package:

```{r eval = FALSE}
install.packages("dplyr")
```

Then load it:

```{r}
library(dplyr)
```

You can see some great, detailed introductions and tutorials in the *vignettes*:

```{r, eval = FALSE}
browseVignettes("dplyr")
```

Cleaning Your Data
------------------

### United Nations voting data

The UN voting data comes from [this page](http://thedata.harvard.edu/dvn/dv/Voeten/faces/study/StudyPage.xhtml?studyId=38311&studyListingIndex=0_dee53f12c760141b21c251525332)- it is otherwise entirely unchanged. You can download it as:

```{r load_data_url, eval = FALSE}
load(url("http://varianceexplained.org/courses/WS1015/files/undata-213.RData"))
```

```{r load_data, echo = FALSE}
load("../files//undata-213.RData")
```

This loads the variables from the RData file into your workspace. What variable is it? You can use `ls()` to find what variables you currently have loaded:

```{r}
ls()
```

*Whatever you do, do not try printing `x`!* It is too big for R to print, and this will crash it! You can, however, display some basic summaries of it:

```{r}
dim(x)
```

You can even view it like a spreadsheet (it won't show the whole thing):

```{r, eval = FALSE}
View(x)
```

The dplyr package provides a way to change x's behavior so that if we accidentally print it, it won't ruin our day. This is the `tbl_df` class:

```{r tbl_df}
x <- tbl_df(x)
class(x)
x
```

Notice that it cuts off after a certain number of columns, and also a certain number of rows. However, it otherwise works just like a data.frame:

```{r}
head(x$rcid)
head(x$session)
```

### Looking at the data's structure

Now, let's look at the code book, which describes each of these columns. It can be downloaded from [this page](http://thedata.harvard.edu/dvn/dv/Voeten/faces/study/StudyPage.xhtml?studyId=38311&studyListingIndex=0_dee53f12c760141b21c251525332). Some of the things it shows are:

* rcid: Roll call vote ID: each of these identifies one vote
* session: One United Nations session: a year
* unres: a UN resolution (there might be multiple votes per resolution)
* vote: Coded vote:
    * 1 = Yes
    * 2 = Abstain
    * 3 = No
    * 8 = Absent
    * 9 = Not a member
* uniqueName: the name of the country

Everything else is lots and lots of other ways of describing countries. For us, these are not important. 

We can find out a bit more about the columns using summary:

```{r}
summary(x)
```

We can see how the roll call ID differs from the UN resolutions (turns out there can be more than one vote per resolution) by using `length` and `unique`:

```{r}
length(unique(x$rcid))
length(unique(x$unres))
```

### dplyr: selecting columns

dplyr provides functions for manipulating our data. We mentioned before that there are some columns we dont care about. dplyr provides the `select` function to extract the columns we want:

```{r}
select(x, rcid, session, date)
select(x, rcid, session, date, unres, vote, uniquename)
```

You can also select multiple consecutive columns using `:`:
 
```{r}
select(x, rcid:uniquename)
select(x, rcid:vote, uniquename)
select(x, rcid:vote, country = uniquename)
```

or can remove specific columns with `-`:

```{r}
select(x, -rcid, -date, -session, -ccode)
```

### The `%>%` operator:

Notice that the first argument to `select` is our data. That is true of all dplyr's functions. If we want to perform multiple operations, this becomes a hassle, because we're nesting function calls within function calls. But dplyr provides another way to write it:

```{r}
x %>% select(rcid:vote, country = uniquename)
```

The `%>%` operator, which is typically pronounced "then", lets us pipe together multiple steps of an analysis. But it's nothing more than a simple conversion:

```{r, eval = FALSE}
a %>% f(b, c)
# becomes
f(a, b, c)
```

```{r, eval = FALSE}
a %>% f(b) %>% g(c, d, e)
# becomes
g(f(a, b), c, d, e)
```

Many data analyses consist of these consecutive operations. This makes the use of `%>%` very natural. So from now on we'll write them like:

```{r, eval = FALSE}
x %>% select(rcid:vote, country = uniquename)
```

### Filter: removing rows based on a condition

Let's say we don't care about the Abstain or Absent votes. We can filter them out using another dplyr function, `filter`:

```{r}
x %>% select(rcid:vote, country = uniquename) %>%
    filter(vote < 8)
```

*Recall* that this is just the same as:

```{r}
filter(select(x, rcid:vote, country = uniquename), vote < 8)
```

But it is already more readable.

### Mutate: changing columns or adding new ones

Right now, votes are represented as 1 (Yes), 2 (Abstain), 3 (No). Let's turn them into a factor.

```{r}
votes <- c("Yes", "Abstain", "No")
x %>% select(rcid:vote, country = uniquename) %>%
    filter(vote < 8) %>%
    mutate(vote = factor(votes[vote]))
```

Secondly, you might be bothered that the country name, and the UN resolution, have quotes around them. The `stringr` package provides a function, `str_replace`, to replace letters in a string with another letter.

```{r, dependson = "tbl_df"}
library(stringr)

x %>% select(rcid:vote, country = uniquename) %>%
    filter(vote < 8) %>%
    mutate(vote = factor(votes[vote]),
           country = str_replace(country, '"', ''))
```


### Dividing date into year/month/day with tidyr's separate

```{r eval = FALSE}
install.packages(tidyr)
```

Right now, year, month and day are combined in the same variable, which limits the operations we can perform with them. Let's try separating them, using dplyr's `separate` operation.

```{r}
library(tidyr)

x %>% select(rcid:vote, country = uniquename) %>% filter(vote < 8) %>%
    mutate(vote = factor(votes[vote]), country = gsub('"', '', country)) %>%
    separate(date, c("year", "month", "day"))
```

Right now, `year`, `month` and `day` are all character vectors. We want them to be numbers. That's handled by the `convert` argument of `separate`. This time, let's save it into a data frame called `votes`:

```{r}
votes <- x %>% select(rcid:vote, country = uniquename) %>%
    filter(vote < 8) %>%
    mutate(vote = factor(votes[vote]), country = gsub('"', '', country)) %>%
    separate(date, c("year", "month", "day"), convert = TRUE)
```

*This* will be the final version of our `votes` data- we've processed the columns and given them reasonable names. Now we can get to the actually interesting operations.

Exploratory Data Analysis
--------------------------

Now that we have the data in the format we want, we can start actually exploring it to answer questions.

### Grouping and Summarizing

An essential operation in data science is the "split-apply-combine" pattern (described [here](TODO)). This breaks up your data into smaller subgroups, performs some analysis on them, and then recombines the results.

This operation by itself doesn't do anything except record, inside the `votes` table, that we're grouping by that variable:

```{r}
votes %>% group_by(year)
```

But when we apply the `summarize` operation later, that operation takes that grouping variable into account, and performs summaries within each `year`:

```{r votesumm_1}
votesumm <- votes %>% group_by(year) %>%
    summarize(numvotes = n())
votesumm
```

Notice there is now one line per year (the original group), containing a new variable, `numvotes`, with the number of votes in that year.

To see why this might be useful, try using `ggplot2` to make a graph of votes per year:

```{r dependson = "votesumm_1"}
library(ggplot2)
ggplot(votesumm, aes(year, numvotes)) + geom_line()
```

You can see that summarizing within each group makes it easy to produce graphs like these. But that's a pretty simple plot. Let's collect more information per year, and plot that. For starters, we could see how the proportion of countries that vote "Yes" on a resolution (a sort of measure of "general agreement") changes from year to year.

```{r votesumm_yes}
votesumm <- votes %>% group_by(year) %>%
    summarize(numvotes = n(), yes = mean(vote == "Yes"))

ggplot(votesumm, aes(year, yes)) + geom_line()
```

Throw in a trend line with `geom_smooth`:

```{r dependson = "votesumm_yes"}
ggplot(votesumm, aes(year, yes)) + geom_line() + geom_smooth()
```

Now, that's about as interesting as we can get while grouping by year. Let's get more interesting and group within *years and countries*. Do this by adding a second variable to the `group_by` 

```{r votesumm}
votesumm <- votes %>% group_by(year, country) %>%
    summarize(numvotes = n(), yes = mean(vote == "Yes"))
```

Notice that there is now one row for each *year-country pair*.

What countries are there to work with? Some of them may have names we're not used to. Here's a trick to visualize them:

```{r}
sort(unique(votesumm$country))
```

Let's grab out a few that might interest us. (Note that the "U.S.S.R." turned into the "Russian Federation" starting in 1992).

```{r}
interesting_countries <- c("United States of America", "U.S.S.R.", "United Kingdom", "Russian Federation")
interesting <- votesumm %>% filter(country %in% interesting_countries)
```

Now that we've filtered for these countries, we can plot their "% Yes" metric separately over time. Here's two ways (of many) you can do this- separating countries by color and by facets (sub-plots):

```{r}
ggplot(interesting, aes(year, yes, color = country)) + geom_point() +
    geom_smooth()
ggplot(interesting, aes(year, yes)) + geom_point() +
    geom_smooth() +
    facet_wrap(~ country)
```

We can already pick out and start interpreting trends based on these four plots, of how each countries level of agreement with the UN's resolutions changed over time.

Next Time
----------

In the next lecture we'll continue diving into this data as an example of a tidy data analysis. We'll learn:

* How to merge this data with a different dataset that describes the topic and importance of each resolution, and create graphs based on these topics
* How to turn un-tidy data into tidy data using [tidyr](https://github.com/hadley/tidyr)
* How to perform an analysis, such as a regression or spline, within each country using the [broom](https://github.com/dgrtwo/broom) package
* How to cluster countries by similarity in voting patterns, and construct heatmaps and trees
