---
layout: post
title: "The adblockr package: block ads from the monetizr package"
description: "Introducing a new package that blocks the greedy, distracting ads
from the heinous monetizr package."
date: 2016-04-01 15:50:00 -0400
category: r
tags: [r]
comments: true
---

I was horrified to learn of the existence of the [monetizr package](http://varianceexplained.org/r/monetizr/), which adds advertisements to R functions. The package goes against the entire philosophy of open source and the spirit of the R community.

Luckily, I was able to construct a fix- the [adblockr package](https://github.com/dgrtwo/adblockr).

### Use freely to get an ad-free experience

For example, imagine someone locked the useful `multiply_by_two` function behind an annoying ad:

```{r}
library(monetizr)

multiply_by_two <- function(x) {
  x * 2
}

multiply_by_two <- marketably(multiply_by_two,
                              "This is an awful, distracting ad")

multiply_by_two(10)
```

The adblockr package offers the `freely` adverb, which converts the `multiply_by_two` function back to give you an ad-free experience:

```{r}
library(adblockr)

multiply_by_two <- freely(multiply_by_two)
multiply_by_two(5)
```

The package also includes the `block_all` function for removing all ads from a package. For example, once the next version of broom is monetized, you can block all ads in the package with:

```{r eval = FALSE}
library(broom)

block_all("broom")
```

If enough of us use this, it will hopefully send a message to the greedy `monetizr` maintainer that ads have no place in R.
